Testing and Isolation Efficacy: Insights from a Simple Epidemic Model

### 2021-05-11 SIR_manuscript.tex
working on discussion, adding a result

Examining our formula for $\Rnum$ \eqref{R0} gives the following results. 
See the appendix, Sec.\,\ref{app:rho} and Sec.\,\ref{app:omega} for details.

@@

\item \label{p1:eta} Increasing the isolation parameters for tested and confirmed individuals decreases \Rnum;

\item \label{p1:rho} Higher testing $\rho$ intensity always decreases $\Rnum$ \textbf{if}: 
%% \begin{enumerate} \item
testing is random; \textbf{or}
%% \item
$\rho$ is small.
%% \end{enumerate}

@@

@@

We also see that greater test intensity $\rho$ increases the effectiveness of control $\Delta$ ($\Delta$ increases as we move up in each plot in this Figure. Mathematically speaking, we did find a counter-vailing effect, but this can only be seen when we allow $\rho$ to take unrealistically large values, and only for weighted testing (see \fref{pan2}). The explanation here is that more rapid testing leaves more susceptibles in the ``waiting-for-no'' category at the DFE; these people must then wait for their tests to be returned before they can be tested again, receive a positive test, and become cautious. This effect is usually weak compared to the helpful effects of testing, but if testing is weighted, $\theta_w$ is low, and test returns are slow, it is possible for increasing $\rho$ to reduce $\Delta$ (see upper-right part of upper-right plot of \fref{pan2}(b)).

@@

{\bf The counter-vailing effect of \percap testing intensity, $\rho$, on $\Rnum$;}
- we are missing out on community-level advantages of fast testing.

{\bf The potential advantage of slow test reporting, or favorable-delay-reporting;}

@@

{\bf The second goal} is to explore how changes in \percap testing intensity $\rho$ affects $\Rnum$. Mathematically we would like to verify the sign of $\pder[\Rnum]{\rho}$, which specifically depends on $\pder[\Delta]{\rho}$. We use the derived expressions for $S\_u^*$ and $S\_n^*$, given by Eqs.\,\eqref{dfe}, in $\Delta$ \eqref{eq:del4}. Also, we define $\phi = \hat F_S = \frac{\rho \omega}{\omega-\rho}$, to reparametrize $\rho$. This is mainly to avoid singularity in $\hat F_I$ \eqref{eq:fi}, when testing intensity $\rho$ is very close to the rate of test return $\omega$. Thus, $\rho$ is reparametrized as 
\begin{equation}
\label{eq:phi}
\rho=\frac{\omega \phi}{\omega+\phi}.
\end{equation}
This one-to-one monotonic reparametrization enables us to simplify the mathematical expressions and explore the simpler $\pder[\Delta]{\phi}$ instead of the complicated $\pder[\Delta]{\rho}$.
The derivative is
\begin{align}
\label{eq:dd3dr}
\partial\Delta/\partial\phi=& \frac{1}{d_3} (a_3\,\phi^2+b_3\,\phi+c_3),
\end{align}
where
\begin{multline}
\label{eq:a3}
a_3 = \frac{w_I}{w_S}
\Bigg(\Big(
( \theta_{{w}}{{\it p\_I}}^{2}-\theta_{{c}}{{\it p\_I}}^
{2}\theta_{{w}} ) {\omega}^{3} 
+ ( 2\,\theta_{{w}}{{\it p\_I
}}^{2}-\theta_{{c}}{{\it p\_I}}^{2}-{\it p\_I}\,{\theta_{{w}}}^{2}-
\theta_{{c}}{\it p\_I}\,\theta_{{w}}-{{\it p\_I}}^{2}{\theta_{{w}}}^{2
}+2\,\theta_{{w}}{\it p\_I} ) \gamma\,{\omega}^{2} \\
 + ( -2\,{ \it p\_I}\,{\theta_{{w}}}^{2}-\theta_{{c}}{\it p\_I}+3\,\theta_{{w}}{
\it p\_I}-{\theta_{{w}}}^{2}+\theta_{{w}} ) {\gamma}^{2}\omega 
 + ( \theta_{{w}}-{\theta_{{w}}}^{2} ) {\gamma}^{3} \Big) \frac{w_I}{w_S}
 + ( -\theta_{{c}}{\it p\_I}\,\theta_{{w}}+\theta_{{c}}{\it p\_I}
 ) \gamma\,{\omega}^{2}+ \\
 ( \theta_{{w}}+\theta_{{c}}{\it
p\_I}-{\theta_{{w}}}^{2}-\theta_{{c}}{\it p\_I}\,\theta_{{w}} )
{\gamma}^{2}\omega+ ( \theta_{{w}}-{\theta_{{w}}}^{2} ) {
\gamma}^{3}
\Bigg),
\end{multline}
\begin{align}
\label{eq:abcd2}
b_3&= 2\frac{w_I}{w_S}(\omega+\gamma)\gamma\Big( 
(\omega+\gamma+\omega\,p_I)(2-\theta\_w)\,\gamma\,\theta\_w+ (1-\theta\_w)\,\omega^2 p_I\theta\_c+\omega^2 p_I\theta\_w 
\Big) ,\\
c_3&= (\omega+\gamma)^2\,\gamma\Big( 
(2-\theta\_w)\,\gamma^2\,\theta\_w+
(1+\frac{w_I}{w_S})\,\omega\,\gamma\,\theta\_w+
\frac{w_I}{w_S} \omega^2 p_I\theta\_c \Big),\\
d_3&= \frac{(\omega+\gamma)}{\omega} \Big((\omega p_I+\gamma)\frac{w_I}{w_S}\phi\,+\,(\omega+\gamma)\gamma \Big)^2(\omega+\phi)^2. 
\end{align}
Note that $\phi\geq 0$, also $b_3$, $c_3$ and $d_3$ are all positive. However $a_3$ can be positive or negative.
If $a_3\geq 0$, $\partial\Delta/\partial\phi \geq 0$ for all feasible range of parameters, thus $\pro\Rnum \leq 0$. 
If $a_3 < 0$, then the quadratic expression in the numerator of \eqref{eq:dd3dr} has a positive root, $\phi^*$, such that for $\phi>\phi^*$, $\partial\Delta/\partial\phi < 0$.

@@

\end{document}

>>>>

Examining our formula for $\Rnum$ \eqref{R0} gives the following results. 
See the appendix, Sec.\,\ref{app:rho}, Sec.\,\ref{app:omega} and Sec.\,\ref{app:w} for details.

@@

\item \label{p1:eta} Increasing the isolation efficacy parameters for tested and confirmed individuals decreases \Rnum;

\item \label{p1:rho} Higher testing intensity $\rho$ always decreases $\Rnum$ \textbf{if}: 
%% \begin{enumerate} \item
testing is random; \textbf{or}
%% \item
$\rho$ is small.
%% \end{enumerate}

@@

\item \label{p1:w} Increasing testing “focus” $\frac{w_I}{w_S}$ from random to highly targeted always decreases \Rnum.

@@

We also see that greater test intensity $\rho$ increases the effectiveness of control $\Delta$ ($\Delta$ increases as we move up in each plot in this Figure). Mathematically speaking, we did find a counter-vailing effect, but this can only be seen when we allow $\rho$ to take unrealistically large values, and only for weighted/targeted testing (see \fref{pan2}). The explanation here is that more rapid testing leaves more susceptibles in the ``waiting-for-no'' category at the DFE; these people must then wait for their tests to be returned before they can be tested again, receive a positive test, and become cautious. This effect is usually weak compared to the helpful effects of testing, but if testing is weighted, $\theta_w$ is low, and test returns are slow, it is possible for increasing $\rho$ to reduce $\Delta$ (see upper-right part of upper-right plot of \fref{pan2}(b)).

@@

{\bf A potential advantage of slow test reporting $\omega$;}
- It seems unapplicable to delay the test return time as a strategy to control the viral spread when $\theta\_w$ is high (moving to the right in each plot as $\theta\_w$ is closer to 1).
-

While targeted testing strategies, including targeting people with infection-like symptoms or the contacts of confirmed cases, are always more effective than random testing, as expected, we find that in some cases the direct effect of testing is that viral spread is greater for a slow test than for a fast test. 
This counter-intuitive effect can occur when people are cautious when awaiting a test result, and may not be robust to second-order effects of fast testing (such as better contact tracing).

{\bf The counter-vailing effect of \percap testing intensity, $\rho$, on $\Rnum$;}
- We are missing out on community-level advantages of fast testing.

{\bf Specifying the compartment-specific relative testing weights, Specifically $\frac{w_I}{w_S}$;}
- More targeted testing is always good in viral control,
- 
We incorporated the compartment-specific relative testing weights, $w_S$, $w_I$ and $w_R$, to model random testing and \emph{targeted} testing strategies. Here, in the case of targeted testing and for the simplicity and illustration purposes, we assumed that infected and recovered individuals are tested at three times the \percap rate of susceptible individuals, thus $w_I/w_S=3$ and $w_R=w_I$. Note that we have not specified a methodology to assign particular relative testing weights corresponding to a particular targeted testing scenario. 
Modeling different targeted testing strategies, equivalently test-specific testing weights in our framework, requires prior information of the conditional probabilities of getting tested for people in a given compartment. 
This can be implied when we would like to quantify and compare the effect of different levels of test focus for infectious people on the basic reproduction number $\Rnum$, and conclude about the disease spread management. For example, when people are tested for "screening", the individuals with potential higher mobility, eg. people who are getting on flights, get more tested and thus the coresponding heavier testing weight is assigned than people awaiting for a surgery and are probably going to stay in a long-term-care facility and consequently less mobile and more isolated to begin with. With our model, we would be able to compare the sensitivity of the disease dynamics, through $\Rnum$, with respect to testing different high-risk groups in the population. This part needs to be developed further in future work.

@@

{\bf The second goal} is to explore how changes in \percap testing intensity $\rho$ affects $\Rnum$. Mathematically we would like to verify the sign of $\pder[\Rnum]{\rho}$, which specifically depends on $\pder[\Delta]{\rho}$. We use the derived expressions for $S\_u^*$ and $S\_n^*$, given by Eqs.\,\eqref{dfe}, in $\Delta$ \eqref{eq:del4}. Also, we define $\phi = \hat F_S = \frac{\rho \omega}{\omega-\rho}$, to reparametrize $\rho$. This is mainly to avoid singularity in $\hat F_I$ \eqref{eq:fi}, when testing intensity $\rho$ is very close to the rate of test return $\omega$. Thus, $\rho$ is reparametrized as 
\begin{equation}
\label{eq:phi}
\rho=\frac{\omega \phi}{\omega+\phi}.
\end{equation}
This one-to-one monotonic reparametrization enables us to simplify the mathematical expressions and explore the simpler $\pder[\Delta]{\phi}$ instead of the complicated $\pder[\Delta]{\rho}$.
The derivative is
\begin{align}
\label{eq:dd3dr}
\partial\Delta/\partial\phi=& \frac{1}{d_3} (a_3\,\phi^2+b_3\,\phi+c_3),
\end{align}
where
\begin{multline}
\label{eq:a3}
a_3 = \frac{w_I}{w_S}
\Bigg(\Big(
( \theta_{{w}}{{\it p\_I}}^{2}-\theta_{{c}}{{\it p\_I}}^
{2}\theta_{{w}} ) {\omega}^{3} 
+ ( 2\,\theta_{{w}}{{\it p\_I
}}^{2}-\theta_{{c}}{{\it p\_I}}^{2}-{\it p\_I}\,{\theta_{{w}}}^{2}-
\theta_{{c}}{\it p\_I}\,\theta_{{w}}-{{\it p\_I}}^{2}{\theta_{{w}}}^{2
}+2\,\theta_{{w}}{\it p\_I} ) \gamma\,{\omega}^{2} \\
 + ( -2\,{ \it p\_I}\,{\theta_{{w}}}^{2}-\theta_{{c}}{\it p\_I}+3\,\theta_{{w}}{
\it p\_I}-{\theta_{{w}}}^{2}+\theta_{{w}} ) {\gamma}^{2}\omega 
 + ( \theta_{{w}}-{\theta_{{w}}}^{2} ) {\gamma}^{3} \Big) \frac{w_I}{w_S}
 + ( -\theta_{{c}}{\it p\_I}\,\theta_{{w}}+\theta_{{c}}{\it p\_I}
 ) \gamma\,{\omega}^{2}+ \\
 ( \theta_{{w}}+\theta_{{c}}{\it
p\_I}-{\theta_{{w}}}^{2}-\theta_{{c}}{\it p\_I}\,\theta_{{w}} )
{\gamma}^{2}\omega+ ( \theta_{{w}}-{\theta_{{w}}}^{2} ) {
\gamma}^{3}
\Bigg),
\end{multline}
\begin{align}
\label{eq:abcd2}
b_3&= 2\frac{w_I}{w_S}(\omega+\gamma)\gamma\Big( 
(\omega+\gamma+\omega\,p_I)(2-\theta\_w)\,\gamma\,\theta\_w+ (1-\theta\_w)\,\omega^2 p_I\theta\_c+\omega^2 p_I\theta\_w 
\Big) ,\\
c_3&= (\omega+\gamma)^2\,\gamma\Big( 
(2-\theta\_w)\,\gamma^2\,\theta\_w+
(1+\frac{w_I}{w_S})\,\omega\,\gamma\,\theta\_w+
\frac{w_I}{w_S} \omega^2 p_I\theta\_c \Big),\\
d_3&= \frac{(\omega+\gamma)}{\omega} \Big((\omega p_I+\gamma)\frac{w_I}{w_S}\phi\,+\,(\omega+\gamma)\gamma \Big)^2(\omega+\phi)^2. 
\end{align}
Note that $\phi\geq 0$, also $b_3$, $c_3$ and $d_3$ are all positive. However $a_3$ can be positive or negative.
If $a_3\geq 0$, $\partial\Delta/\partial\phi \geq 0$ for all feasible range of parameters, thus $\pro\Rnum \leq 0$. It is straight forward to show that $a_3\geq 0$ in case of random testing strategy, i.e., $w\_S=w\_I=1$. 
If $a_3 < 0$, then the quadratic expression in the numerator of \eqref{eq:dd3dr} has a positive root, $\phi^*$, such that for $\phi>\phi^*$, $\partial\Delta/\partial\phi < 0$.

@@

% %%%%%%%
\subsection{The effect of testing focus parameter $\frac{w_I}{w_S}$ on $\Rnum$} \label{app:w}

We define $w\_{IS}=\frac{w_I}{w_S}$.
\begin{align}
\pder[\Delta]{w\_{IS}}= \dfrac{(\omega-\rho)\,(\omega(\omega-\rho\theta\_w)+\gamma(\omega-\rho))\,(\theta\_w\gamma+\theta\_c\omega\,p\_I)}
{(-\omega^2\,\gamma+\omega\,\gamma\,\rho-\gamma\,\rho\,\omega\,w\_{IS}-
\omega\,\gamma^2+\gamma^2\,\rho-\omega^2\,p_I\,\rho\,w\_{IS})^2},
\end{align}
which is a positive quantity. Thus, $\pder[\Rnum]{w\_{IS}}\leq 0$. Therefore, increasing the focus of testing on the infectious people will result in less transmission.  
\end{document}
